\textbf{Dynamic Hover Detection:\\Creating GUIs in Elm}

### 2013-12-15 paper/paper.tex
More revisions

Section 2 introduces Elm, signals, and the prohibition on signals of signals. It
is targeted to readers familiar with functional programming but not FRP, and may
be skipped by those already familiar with Elm. Section 3 presents Elm's hover
detection for DOM elements, why it initially appears inadequate, and how it can
be extended to meet our needs. Section 4 explains our menu implementation in
detail. Section 5 analyzes TodoFRP. Section 6 concludes with a notice to the Elm
community.

@@

Elm provides a function to obtain the hover information from an Element.

@@

The returned Element is visually identical to argument, but now detects
hover information. The signals of Booleans reflects the hover status of
the this Element, not the original. This function works well when the
Element is pure (not a signal). For example, if the top-level menu items are
known statically and never change, they can be rendered as pure Elements. The
\texttt{hoverable} function allows us to create Elements for their submenus.
However, if we try to lift the function, we get

@@

Although it may not appear to be so at first, the result type is a signal of
signals. It is possible to transform the result into \texttt{Signal Signal
(Element, Bool)}. A more mentally convenient type, though impossible to obtain,
is \texttt{(Signal Element, Signal Signal Bool)}. There does not exist a general
join function to operate on the \texttt{Signal Signal Bool}.

This is rather unfortunate, as we need to detect hover information on a dynamic
element for a number of reasons. First, after a menu is extended by hovering
over its parent, it must remain so as long as the mouse hovers over the menu
itself. Secondly, if we wish the indicate the currently hovered-over item to the
user by visually highlighting it, the element must be dynamic. Thirdly, we do
not wish to require that menu labels be string literals in the source code,
which would be necessary if we could not use signals of elements. Rather they
may be retrieved or generated in some other manner, or they may change as the
user switches among applications. Finally, in order to determine when a user has
clicked on an Element so that the system may respond, we must know which one is
being hovered over.\footnote{Once this is known, \texttt{sampleOn Mouse.clicks}
can be applied to obtain a signal of the hovered-over element updated on each
click.}

@@

It is done using the more general \texttt{hoverables} (note the plural)
primitive, of the following type:

@@

It is not just hover detection that follows the convention of the two
names primitives, singular and plural, but most of Elm's wrappers around
HTML GUI components. They are therefore receptive of the same join
technique. For labeled buttons, press events are represented by unit or
a provided identifier:

@@

Drop-down menus do not use the same paradigm as the rest; there is
\texttt{dropDown} but no \texttt{dropDowns}. This is unfortunate not
only because there seems to be no way to supply a
\texttt{Signal {[}String{]}}, but because drop-downs are commonly used
in groups. Multiple menus reporting on one signal would be very
convenient. If the menus were dynamic, it would be useful to determine
when an event occurred because the user makes a selection vs.~when the
menu updated and the previously selected string was no longer available.

% These are likely unnecessary and cause page breaks in bad places.
%
%\texttt{Graphics.Input} does not currently support radio buttons.
%
%As an aside, \texttt{button} is an impure function. One can create two
%buttons with the same label and place them in different places on the
%screen (or place one not at all), and the signals of press events will
%be different. Lifting a joined version onto a signal of strings does not
%change this fact.

>>>>

Section 2 introduces Elm, signals, state, and the prohibition on signals of
signals. It is targeted to readers familiar with functional programming but not
FRP, and may be skipped by those already comfortable with Elm. Section 3
presents Elm's hover detection for DOM elements, why it initially appears
inadequate, and how it can be extended to meet our needs. Section 4 explains our
menu implementation in detail. Section 5 analyzes TodoFRP. Section 6 concludes
with a notice to the Elm community.

@@

Elm's \texttt{Graphics.Input} library provides a function to obtain the hover
information from an Element.

@@

The returned Element is visually identical to argument, but now detects
hover information. The signal of Booleans reflects the hover status of
the this Element, not the original. This function works well when the
Element is pure (not a signal). For example, if the top-level menu items are
known statically and never change, they can be rendered as pure Elements with
this function. We can also create a function, perhaps of type \texttt{Bool ->
Element}, which we can lift onto the hover data to display the submenu only on
mouseover. The result is a value of type \texttt{Signal Element}. If we try to
lift \texttt{hoverable} to accept this value, we get

@@

Although it may not appear to be so at first, the result type is a signal of
signals. It is possible to transform the result into \texttt{Signal Signal
(Element, Bool)}. A more mentally convenient type, though impossible to
obtain,\footnote{We explain why in Section 4.}
is \texttt{(Signal Element, Signal Signal Bool)}. There does not exist a general
join function to operate on the \texttt{Signal Signal Bool}.

This is rather unfortunate, as we need to detect hover information on a dynamic
element for a number of reasons. First, after a menu is extended by hovering
over its parent, it must remain so as long as the mouse hovers over the menu
itself. Without this information, the menu will disappear as soon as we try to
mouse over it. Secondly, if we wish the indicate the currently hovered-over item
to the user by visually highlighting it, the element must be dynamic. Thirdly,
we do not wish to require that menu labels be string literals in the source
code, which would be necessary if we could not use signals of elements. Rather,
they may be retrieved or generated in some other manner, or they may change as
the user switches among applications. Finally, in order to determine when a user
has clicked on an Element so that the system may respond, we must know which one
is being hovered over.\footnote{Once this is known, \texttt{sampleOn
Mouse.clicks} can be applied to obtain a signal of the hovered-over element
updated on each click.}

@@

Comparing its type to the mentally convenient type of \texttt{lift hoverables},
we see it is identical except that the Boolean value is a signal, not a signal
of signals. This function is implemented using the more general
\texttt{hoverables} (note the plural) primitive, of the following type:

@@

It is not just hover detection that follows the paradigm of two primitives
with singular and plural names. Most of Elm's wrappers around HTML GUI
components do so as well, and have analogous types. They are therefore receptive
of the same join technique. For text-labeled buttons, press events are
represented by unit or a provided identifier:

@@

Drop-down menus do not use the same paradigm as the rest; there is
\texttt{dropDown} but no \texttt{dropDowns}. This is unfortunate not
only because there seems to be no way to supply a
\texttt{Signal {[}String{]}}, but because drop-downs are commonly used
in groups. Multiple menus reporting on one signal would be very
convenient. If the menus were dynamic, it would be useful for the API to
indicate when an event occurred because the user makes a selection vs.~when the
menu updated and the previously selected string was no longer available.

\texttt{Graphics.Input} does not currently support radio buttons, but one could
imagine them being added in a way similar to the others. However, it is worth
observing that the pattern itself leaves something to be desired, namely, that
the functions are impure. One can apply \texttt{hoverable} to an Element twice,
place the resulting Elements in in different places on the screen (or place one
not at all), and the signals of Booleans will be different. This is true,
analogously, of all the functions of similar type in the library. We know of no
alternative that has been proposed, and do not have one ourselves. However, we
point out this deficiency in order to illustrate that the library still has room
for incremental and perhaps even radical growth.
